Project 7 Healthcare cost analysis


```{r}
install.packages(gdata)
library (gdata)
health=read.csv("1555054100_hospitalcosts - HospitalCosts.csv")
health
```

```{r}
str(health)
summary(health)

```


```{r}
#health$FEMALE = as.factor(health$FEMALE)
#health$AGE = as.factor(health$AGE)
#health$RACE = as.factor(health$RACE)

#str(health)
```


1. To record the patient statistics, the agency wants to find the age category of people who frequently visit the hospital and has the maximum expenditure.



```{r}
hist(health$AGE)

library(dplyr)

arrange(group_by(health, AGE) %>%
  summarise(count=n(),
            sum=sum(TOTCHG)), desc(sum))
```

```{r}

healthage = as.factor(health$AGE) 
  which.max(summary(healthage))
max(summary(healthage))

which.max(tapply(health$TOTCHG,health$AGE,sum))
max(tapply(health$TOTCHG,health$AGE,sum))

```

The age "0" (infant) category of children is a group who frequently visit the hospital (307 times).
Max expenditures are in the group 0 (infant)- $678.118.


```{r}

library(ggplot2)
ggplot(health, aes(x=healthage))+geom_bar()


summary(healthage)
```




2. In order of severity of the diagnosis and treatments and to find out the expensive treatments, the agency wants to find the diagnosis-related group that has maximum hospitalization and expenditure.


```{r}
arrange(group_by(health, APRDRG) %>%
  summarise(count=n(),
            sum=sum(TOTCHG)), desc(sum))
```



```{r}

which.max(summary(as.factor(health$APRDRG)))
max(summary(as.factor(health$APRDRG)))
#summary(as.factor(health$APRDRG))
which.max(tapply(health$TOTCHG,health$APRDRG,sum))
max(tapply(health$TOTCHG,health$APRDRG,sum))
```


```{r}
attach(health)

hist(APRDRG)
```

```{r}

#group1=as.factor(health$APRDRG)
#health = mutate(health, TOTCHG1 = TOTCHG/1000)
#table2 = tapply(health$TOTCHG1,group1,sum)
#barplot(table2)
```

3. To make sure that there is no malpractice, the agency needs to analyze if the race of the patient is related to the hospitalization costs.


```{r}
ggplot(health,aes(x=RACE, y=TOTCHG))+geom_point()
ggplot(health,aes(x=RACE))+geom_bar()
healthrace = as.factor(health$RACE)
summary(healthrace)
```

```{r}

health1 = na.omit(health)
health_anova=aov(TOTCHG~as.factor(RACE), data=health1)
summary(health_anova)

```


```{r}
stripchart(TOTCHG~RACE, vertical = T, pch = 19, data = health1)
analysis=lm(TOTCHG~RACE, data=health1)
summary(analysis)
anova(analysis)
plot(analysis, which=2)
sresids=rstandard(analysis)
hist(sresids)


```



In H0 we assume that all patients from different race groups are treated equally and the same treatment and hospital costs are applied.
Therefore:
H0 - the means of the 6 race group are the equal. 
H1  - not all of the means are equal

P-value 0.943 is greater than 0.05 significant level so H0 can not be rejected. 
It means that there is no relations between race of patient and the hospital costs.
From the summary we can see that RACE 1 has the maximum patients 484 out 500. Also Normal Q-Q chart shows that our data has normal distribution on the beginning and at the end is very much skewed. It effects also Anova results.


4. To properly utilize the costs, the agency has to analyze the severity of the hospital costs by age and gender for the proper allocation of resources.

```{r}

healthfemale = as.factor(health$FEMALE)
ggplot(health,aes(x=healthfemale, y=TOTCHG))+geom_point()
ggplot(health,aes(x=healthfemale))+geom_bar()
  
arrange(group_by(health, FEMALE) %>%
  summarise(count=n(),
            sum=sum(TOTCHG)), desc(sum))

model1 = lm(TOTCHG~AGE + FEMALE, data=health1)
summary(model1)
```


Age has a significant level and lowest p-value that means that Age has significant impact on hospital costs. It seems there is a  relationship between gender and costs too as Female has less p-value. Negative co-efficient in Female shows that female patients uncur less less costs than male. We can prove it also from the summary table. 


5. Since the length of stay is the crucial factor for inpatients, the agency wants to find if the length of stay can be predicted from age, gender, and race.


```{r}
hist(health$LOS)
boxplot(health$LOS)

model2 = lm(LOS~AGE + FEMALE + RACE, data=health1)
summary(model2)

```

The very high p-value means that there is no relationship between independent and dependent variables. Therefore we cannot predict  the length of stay from age, gender and race.

6. To perform a complete analysis, the agency wants to find the variable that mainly affects hospital costs.

```{r}

model3 = lm(TOTCHG~ ., data=health1)
summary(model3)
```




